A supertree pipeline for summarizing phylogenetic and taxonomic information for millions of species

### 2016-10-11 doc/paper1-submission/paper4.tex
Update.

\caption{Representation of input splits in the \otl{} tree and the \otlprop{} tree.}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
 & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolves{} & \resolvedby{}\tabularnewline
\hline 
\otl{} & 34,595 & 923 & 45,385 & 2760 & 473 & 2718\tabularnewline
\hline 
\otl{} only & 745 & 54 & 0 & 2055 & 0 & 2718\tabularnewline
\hline 
\otlprop{} & 38,521 & 1,118 & 45,385 & 1,357 & 515 & 0\tabularnewline
\hline 
\otlprop{} only & 4,671 & 249 & 0 & 652 & 42 & 0\tabularnewline
\hline
\label{tab:representinputs}
\end{tabular}}

@@

\caption{Representation of taxonomy splits in the \otl{} tree and the \otlprop{} tree.}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
 & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolves{} & \resolvedby{}\tabularnewline
\hline 
\otl{} & 125,384 & 0 & 2,424,255 & 1998 & 3676 & 4\tabularnewline
\hline 
\otl{} only & 296 & 0 & 0 & 19 & 17 & 4\tabularnewline
\hline 
\otlprop{} & 125,107 & 0 & 2,424,255 & 2,279 & 3,883 & 0\tabularnewline
\hline 
\otlprop{} only & 19 & 0 & 0 & 300 & 224 & 0\tabularnewline
\hline 
\label{tab:representtaxonomy}
\end{tabular}}

@@

We performed an annotation of both the \otl{} tree and the \otlprop{}
tree as described in section \ref{sec:Annotation} to assess the
ability of our new supertree method to represent splits from input
phylogenies (Table \ref{tab:representinputs}).  This annotation
classifies input phylogeny splits according to how they relate to a
summmary tree, so that each input edge falls in one of \supportedby{},
\partialpathof{}, \terminal{}, \conflictswith{}, or \resolvedby{}.

The number of displayed input splits (supported\_by +
partial\_path\_of) increased from 35,518 (for version 4) to 39,639; an
11.2\% increase. The \otlprop{} tree displays 4920 input splits that
are not displayed by the \otl{} tree, but does not display 799 input
splits that are displayed by the \otl{} tree.

Unlike the version 4 tree, the tree from propinquity contained no
unsupported nodes or unnecessary polytomies. The number of input
splits that do not conflict with the summary tree, but are not
incorporated (resolved\_by) dropped from 2718 to 0.

The number of input splits that conflict with the summary
(conflicts\_with) dropped from 2760 to 1357, a decrease of 1357, or
49.2\%.

@@

Here we have described the motivation and methodology used by the
propinquity tool that is currently used by the Open Tree of Life project
to build its summary tree. A modified version of the treemachine software
which built the summary tree described in the \citet{HinchliffEtAl2015}
paper is used by the project to serve the tree produced by propinquity
via Open Tree of Life APIs. The migration of summary tree construction
from treemachine (used for version 4) to propinquity (for all versions
from v5.0 to present) has decreased the computational time required
to construct a supertree from several hours to about 8 minutes (after
some preprocessing steps which only have to be performed when the
input taxonomy changes). The amount of RAM required during tree construction
has also decreased substantially.

>>>>

\caption{Representation of input splits in the \otl{} tree and the \otlprop{} tree.}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
 & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolvedby{} & \resolves{}\tabularnewline
\hline 
\otl{} & 34,595 & 923 & 45,385 & 2,760 & 2,718 & 473\tabularnewline
\hline 
\otl{} only & 745 & 54 & 0 & 2,055 & 2,718 & 0\tabularnewline
\hline 
\otlprop{} & 38,521 & 1,118 & 45,385 & 1,357 & 0 & 515\tabularnewline
\hline 
\otlprop{} only & 4,671 & 249 & 0 & 652 & 0 & 42\tabularnewline
\hline
\label{tab:representinputs}
\end{tabular}}

@@

\caption{Representation of taxonomy splits in the \otl{} tree and the \otlprop{} tree.}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
 & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolvedby{} & \resolves{}\tabularnewline
\hline 
\otl{} & 125,384 & 0 & 2,424,255 & 1998 & 4 & 3676\tabularnewline
\hline 
\otl{} only & 296 & 0 & 0 & 19 & 4 & 17\tabularnewline
\hline 
\otlprop{} & 125,107 & 0 & 2,424,255 & 2,279 & 0 & 3,883\tabularnewline
\hline 
\otlprop{} only & 19 & 0 & 0 & 300 & 0 & 224\tabularnewline
\hline 
\label{tab:representtaxonomy}
\end{tabular}}

@@

We performed an annotation of both the \otl{} tree and the \otlprop{}
tree as described in section \ref{sec:Annotation} to assess the
ability of our new supertree method to represent splits from input
phylogenies.  Table \ref{tab:representinputs} classifies the input
phylogeny splits according to how they relate to a summmary tree, so
that each input edge falls in one of \supportedby{}, \partialpathof{},
\terminal{}, \conflictswith{}, or \resolvedby{}. For example, the
numbers in the \conflictswith{} column indicate the number of input
splits $j$ with at least one summary edge $x$ such that the relation
``$x$ \conflictswith{} $j$'' holds.  The total number of non-terminal
input phylogeny splits considered thus 40996.

The number of displayed input splits (\supportedby{}
+ \partialpathof{}) increased from 35,518 (for \otl{}) to 39,639 (for \otlprop{});
an 11.2\% increase. When examining which splits are displayed, we find
that the \otlprop{} tree displays 4920 input splits that
are not displayed by the \otl{} tree, whereas the \otl{} tree displays
only 799 input splits that are not displayed by the \otlprop{} tree.  The number of input
splits that conflict with the summary (conflicts\_with) dropped from
2760 to 1357, a decrease of 1357, or 49.2\%.  In accordance with the
goal of not containing unnecessary polytomies, the number of input
splits that do not conflict with the summary tree, but are not
incorporated (resolved\_by) dropped from 2718 to 0.  We also find that
the number of polytomies in input phylogenies that are resolved by the
summary tree increases from 473 for \otl{} to 515 for \otlprop{}.

We also performed an annotation of the \otl{} tree and the \otlprop{}
tree to assess the degree to which these trees represent taxonomy
splits (Table \ref{tab:representtaxonomy}).  The \otlprop{} tree
conflicts with 281 more taxonomy splits than the \otl{} tree.  Since
the taxonomy is the lowest ranked input tree, this increased conflict
with the taxonomy is an expected result of incorporating more splits
from input phylogenies.

@@

Here we have described the motivation and methodology used by our new
supertree method that is currently used by the Open Tree of Life
project to build its summary tree.  Our new method represented 11\%
more input phylogeny splits with 49\% less conflict compared to the
Open Tree of Life version 4 summary tree when applied to the same inputs.
Unless the previous method \citep{HinchliffEtAl2015}, our new method
is guaranteed to incorporate input splits that do not conflict with
the summary tree. The method is implemented in the Open Source
software package propinquity.  A modified version of the treemachine
software which built the summary tree described in the
\citet{HinchliffEtAl2015} paper is used by the project to serve the
tree produced by propinquity via Open Tree of Life APIs.

The migration of summary tree construction from treemachine (used for
version 4) to propinquity (for all versions from v5.0 to present) has
increased the pace of synthesis tree releases from the Open Tree of
Life project.  This is partly because the newly available annotations
feature has made it possible to identify which input trees are
responsible for taxa being included or excluded from the summary tree.
Additionally, the new propinquity software pipeline has decreased the decreased
the computational time required to construct a supertree from several
hours to about 8 minutes (after some preprocessing steps which only
have to be performed when the input taxonomy changes).  The amount of
RAM required during tree construction has also decreased substantially.
